Race-based disparities in allocation of academic disciplinary actions are associated

### 2018-03-13 written_docs/draft.Rmd
Intro! plus rewrite of results & methods

---
title             : "Race-based disparities in allocation of academic disciplinary actions are associated with county-level rates of bias"
shorttitle        : "Discipline Disparities"

author: 
  - name          : "Travis Riddle"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "triddle@princeton.edu"
  - name          : "Stacey Sinclair"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Princeton University"

author_note: >

  Enter author note here.

abstract: >
  Enter abstract here.

keywords          : "keywords"
wordcount         : "X"

bibliography      : ["ref.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
output            : papaja::apa6_word
---

```{r include = FALSE}
library("papaja")
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, cache.lazy = FALSE, 
                      message=FALSE, warning=FALSE, results='hide')
```
# Introduction

# Methods
## Data Sources
```{r summary-data}
library(dplyr)
library(forcats)
library(haven)
library(tidyr)
county_means <- read.csv('/Users/travis/Documents/gits/educational_disparities/output/county_means.csv', 
                         colClasses = 'character')
schools_dat <- read.csv('/Users/travis/Documents/gits/Data/crdc201314csv/CRDC2013_14_SCH.csv')
model.data <- read.csv('/Users/travis/Documents/gits/educational_disparities/output/full_model_data.csv')

@@

```
We used three distinct data sources for the work described here. The academic disciplinary data is part of the Civil Rights Data Collection (CRDC) through the US Department of Education. The dataset we used comes from the 2013-2014 academic year and has data on "all public local and educational agencies and schools, including long-term secure juvenile justice facilities, charter schools, alternative schools, and schools serving students with disabilities." In total, the CRDC data represents `r dim(schools_dat)[1]` institutions enrolling approximately 50 million students, of which approximately 25 million are white and 7.8 million are Black[^1]. Previous work using these data have identified a number of districts whose data are in error, and have excluded juvenile justice facilities, as these institutions constitute dramatically different educational environments, where the meaning of disciplinary actions may be quite different [@losen2015we]. After these exclusions are applied, the final sample used for modeling consists of `r enrollment_totals$total_schools[1]` institutions, enrolling `r round(sum(enrollment_totals$ns)/1000000, 1)` million black or white students, of which `r  round(enrollment_totals$ns[2]/1000000, 1)` million are white and `r  round(enrollment_totals$ns[1]/1000000, 1)` million are black.

@@

We used measurements of implicit and explicit bias available from data collected through Project Implicit [@xu2014psychology]. From this data, we used only respondents who had geographic information that would allow us to place them in a United States county, identified as White, and visited the site before 2015. This consisted of approximately 1.1 million total respondents from `r dim(county_means)[1]` counties.

```{r teachers-summary}

>>>>

---
title: Race-based disparities in allocation of academic disciplinary actions are associated
  with county-level rates of bias
author:
- address: ''
  affiliation: '1'
  corresponding: yes
  email: triddle@princeton.edu
  name: Travis Riddle
- affiliation: '1'
  name: Stacey Sinclair
affiliation:
- id: '1'
  institution: Princeton University
output:
  papaja::apa6_word: default
  papaja::apa6_pdf: default
bibliography: ref.bib
class: man
figsintext: yes
figurelist: no
footnotelist: no
keywords: keywords
lang: english
lineno: yes
author_note: |
  Enter author note here.
shorttitle: Discipline Disparities
tablelist: no
abstract: |
  Enter abstract here.
wordcount: X
---

```{r include = FALSE}
library("papaja")
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, cache.lazy = FALSE, 
                      message=FALSE, warning=FALSE, results='hide')
```
# Introduction
In comparison to White Americans, African Americans exhibit poorer educational outcomes across a range of metrics. One outcome of particular concern is the gap in disciplinary actions [@kinsler2011understanding; @okonofua2016vicious]. Research using administrative datasets and longitudinal samples clearly show that African American students are far more likely to be suspended or expelled [@aud2011america; @yeager2017loss], and conditional on an office referral, are more likely to receive stiffer punishments [@gregory1995crime]. Concerns about these disparities are exacerbated by their associations with long-term life outcomes, including prospects for employment [@pager2009sequencing], mental and physical health [@pascoe2009perceived], and involvement in the criminal justice system [@hirschfield2009another]. In this paper, we aim to establish that these disciplinary gaps are related to environmental levels of bias in predictable ways.

It is perhaps not surprising that levels of bias would be associated with racial disciplinary gaps. This straightforward statement, however, is contextualized in a number of ways. First, the operationalization of bias should play a role in the association. As we describe below, levels of implicit and explicit biases should be differentially associated with disciplinary actions. This differential association should be determined, in part, by the type of disciplinary action under examination. For a given student committing a specific offense, the educational institution may opt for one of several disciplinary actions. These actions vary in severity, institutional costs, and the degree of formal administrative involvement.

To illustrate this problem in the context of previous research, we now describe research on the associations between implicit bias and explicit bias and decision making. This sets the stage for understanding what types of decisions administrators and educators make when electing a specific disciplinary action. We then describe the types of disciplinary actions that are taken and the characteristics of these actions in terms of severity, institutional costs, and formal administrative involvement. Finally, in crossing this information, we concretely specify the set of associations expected when examining bias and disciplinary actions in American schools.

## Bias and decision-making
Previous research on biased attitudes and behavior has shown that although implicit and explicit attitudes are typically correlated [@mcconnell2001relations; @hofmann2005meta], they tend to behaviorally manifest slightly differently. As a default, a naiive expectation is that explicit attitudes should guide behavior, since if people tell us they like something, we should expect that to generally correspond to their behavior. However, under some circumstances, implicit biases are more strongly associated with behavioral outcomes. @nosek2011implicit suggest that, in comparison to explicit attitudes, implicit attitudes are more strongly associated with behavior when the individual lacks motivated to explicitly respond differently than their implicit attitude would suggest, when they do not have the opportunity or ability to modify their initial response inclination, or when they do not have awareness of the implicit response, the behavior, or the link between them. In contrast, the more motivated, more able, or more aware the person is over their behavioral response, the more explicit biases should play a role. One key situational factor that determines the response is situational uncertainty. In uncertain situations, the respondent is not likely to have strong motivations to change their initial response, otherwise the situation would not be uncertain. Research supports this idea, as when the situation is characterized by greater uncertainty, implicit biases are known to be stronger determinants of behavior [@friese2009and; @gawronski2003implicit; @galdi2008automatic]. In short, when taking (or not taking) an action can be attributed to multiple reasons due to situational ambiguity, then implicit biases should result in higher rates of discriminatory behavior.

Accordingly, this research suggests that disciplinary actions that are implemented quickly, have relatively little administrative oversight, and require little in the way of insitutional resources should be characterized by strong associations with implicit bias. On the other hand, any disciplinary action that requires higher levels of oversight, resources, or lengthy time investments to complete should be more strongly associated with explicit bias.

## Types of disciplinary actions
We report here on five different types of disciplinary actions: in-school suspensions, out-of-school suspensions, law enforcement referrals, school-related arrests, and expulsions. As described above, these actions vary in severity, required resources, and administrative oversight.

The most extreme outcome, expulsion, is defined as when a student is prohibited from ever returning to the educational institution. The institution may or may not set up alternative educational services for the student. Because of the the severe nature of this punishment, it also requires considerable administrative oversight, and is often conferred for behaviors that are relatively unambiguous (e.g. bringing a gun to school). We believe similar cases can be made for law enforcement referrals and school-related arrests insofar as these actions tend to be more severe, and require resources, or occasionally the involvement of outside agencies.

Somewhat less severe than expulsions, but one of the most frequently utilized disciplinary actions, out-of-school suspensions occupy something of a middle ground. There is a substantial body of research documenting the negative associations between this particular disciplinary action and life outcome [@cuellar2015school; @matjasko2011effective], as well as their disproportionate use on stigmatized students [@haight2014ecological; @kinsler2011understanding]. Out-of-school suspensions are administered for a wide range of behaviors. Some of these behaviors are apt to be determined with a fair degree of subjectivity (e.g. defiance) [@gregory2008discipline], while others may be more unambiguous (e.g. fighting, drug use). In comparison to expulsions and arrests, this type of action demands less administrative oversight.

In-school suspensions, wherein the student is removed from the classroom but remains under the direct supervision of school personnel, are also administered for a wide range of behaviors. However, in-school suspensions have been highlighted as a promising alternative to more exclusionary types of disciplinary actions (e.g. out-of-school suspensions, expulsions) partly because students remain under the supervision of school personnel, which keeps them in a structured, educational setting [@anyon2014persistent]. Like out-of-school suspensions, this type of action requires relatively little administrative oversight. But unlike out-of-school suspension, removing a student from a classroom to a *different* educational setting requires additional space and personnel, and so this represents an inherent investment in the student.

## Concrete predictions of association
Given these disciplinary outcomes and what is known about implicit and explicit attitudes, what should we expect in examining associations between them? First, it's clear that we should expect African American students to receive higher levels of disciplinary actions. Second, we should expect the disparity in disciplinary actions to generally be positively associated with explicit bias. Finally, we should expect implicit bias to only show associations in instances where there is little administrative overhead, and where the behaviors to-be-punished are often ambiguous. By this logic, out-of-school suspensions are the only metric that meet these criteria.

## The present study
The present study uses a national dataset of disciplinary outcomes to evaluate the degree to which implicit and explicit bias are associated with each of these outcomes. After an initial, registered analytic approach, we subsequently made several adjustments to strengthen our analyses and streamlined them with respect to previous research. In general, the data support the conclusion that implicit biases are specifically associated with out-of-school suspensions, while explicit biases tend to be associated with disciplinary actions of all types.

# Methods
## Data Sources
### Disciplinary actions
```{r summary-data}
library(dplyr)
library(forcats)
library(haven)
library(tidyr)
library(ggplot2)
county_means <- read.csv('/Users/travis/Documents/gits/educational_disparities/output/county_means.csv')
schools_dat <- read.csv('/Users/travis/Documents/gits/Data/crdc201314csv/CRDC2013_14_SCH.csv')
model.data <- read.csv('/Users/travis/Documents/gits/educational_disparities/output/full_model_data.csv')

@@

```
To assess rates of disciplinary action, we used data from the Civil Rights Data Collection (CRDC) conducted by the US Department of Education. The dataset we used comes from the 2013-2014 academic year and has data on "all public local and educational agencies and schools, including long-term secure juvenile justice facilities, charter schools, alternative schools, and schools serving students with disabilities." In total, the CRDC data represents `r dim(schools_dat)[1]` institutions enrolling approximately 50 million students, of which approximately 25 million are white and 7.8 million are Black[^1]. Previous work using these data have identified a number of districts whose data are in error, and have excluded juvenile justice facilities, as these institutions constitute dramatically different educational environments, where the meaning of disciplinary actions may be quite different [@losen2015we]. After these exclusions are applied, the final sample used for modeling consists of `r enrollment_totals$total_schools[1]` institutions, enrolling `r round(sum(enrollment_totals$ns)/1000000, 1)` million black or white students, of which `r  round(enrollment_totals$ns[2]/1000000,1)` million are white and `r  round(enrollment_totals$ns[1]/1000000, 1)` million are black. From these data, we focus on the number of students by race (black and white) who received one of several types of disciplinary action. We report here rates of out-of-school suspension, in-school suspension, school-related arrests, law enforcement referrals, and total number of expulsions of any type.

@@

We used measurements of implicit and explicit bias available from data collected through Project Implicit [@xu2014psychology]. For a full description of the implicit and explicit bias measures available in these data, refer to @xu2014psychology and @leitner2016blacks. We note only that we used the IAT D-score as a measure of implicit bias, and the difference between reported warmth towards whites and warmth towards black (both measured from 0=very cold, 10=very warm) as a measure of explicit bias. Additionally, we used only respondents who had geographic information that would allow us to place them in a United States county, identified as White, and visited the site before 2015. This consisted of approximately 1.1 million total respondents from `r dim(county_means)[1]` counties.

### Covariates
Each county-level variable used as a covariates in the final model and the corresponding state-level variable used as a predictor in the post-stratification scheme were taken from the from the same source. Population size and proportions, socioeconomic indicators, and mobility were all taken from the American Community Survey (ACS) 5-year estimates for the time period ending in 2014. Urban-rural indicators were taken from the 2010 US Census, and crime rates were taken from the FBI Uniform Crime Reporting program, as made available through the National Archive of Criminal Justice Data for each year from 2010-2014. Each of these variables is described below.

#### Population size and proportions
We obtained the total population, the proportion of the population that is white, the proportion of the population that is black, and the ratio of black-to-white people in the population from ACS table B02001.

#### Socioeconomic indicators
We obtained estimates for the percentage of population with a Bachelor's degree or higher, the percentage of the population aged 16 or over in the labor force that is unemployed, the median household income, and the percentage of families and people whose income in the last year was below the poverty line from the ACS table DP03.

#### Urban-rural indicator
We obtained estimates of housing density per square mile of land area from Census table GCT-PH1.

#### Mobility
We obtained estimates of population mobility by summing the percentage of African Americans who moved from a different county, state, or country into the county of interest (county-level covariate) or who moved from a different state or country into the state of interest (state-level covariate). We took these metrics from the ACS table S0701.

#### Crime
We computed estimates of the number of violent crimes per person by taking the number of crimes reported divided by the population size for each year, and averaging the resulting proportions across the 5 years of data.

## Data analysis
Because Project Implicit is a nonrandom sample, we used multilevel regression and post-stratification to obtain accurate geographical population-based estimates of implicit and explicit bias. Multilevel regression and post-stratification corrects for biased sampling and regularizes extreme observations with little data to support them (e.g. a county with only a handful of respondents with especially high or low scores) [@park2004bayesian; @gelman1997poststratification]. Following past work [@leitner2016blacks], we identified age as one dimension along which IAT respondents differed from the general population in ways that could bias our conclusions [@gonsalkorale2009aging]. Our post-stratification weighting scheme is as follows: We first grouped respondents into five age group categories (15-24, 25-34, 35-54, 55-75, and 75+). We next fit multilevel models estimating bias (implicit and explicit biases seperately) as a function of our state-level covariates (the "fixed" effects: proportion of population that is white; proportion of the population that is black, the ratio of black to white people in the population, the percentage of college graduates, the unemployment rate, the median household income, the poverty rate, housing density, mobility rate, & crime rate; each of these is described in more detail below), and allowed the estimates to vary by age bin, county, and state (the "random" effects). Next, we determined the population of whites in each county in these age groups using the American Community Survey's 5-year estimates ending in 2014. Finally, we used our estimated models to predict the expected response for each age bin, in each county. Our final county-level estimates are the average of the values predicted for the 5 age bins, weighted by the population size of that bin in that county. As a result of this procedure, we can be confident that our county-level estimates should more closely approximate what our estimates would look like if the Project Implicit data were truly representative along the age dimension in all counties.

We analyzed these data using a series of bayesian multilevel logistic regressions. We modeled the probability that a student would be expelled as a function of a set of effects that are constant across observations (i.e. fixed effects: race (dummy coded), implicit bias, explicit bias, an interaction term between race and implicit bias, an interaction term between race and explicit bias, and all covariates described above) and a set of effects that vary across counties (i.e. random effects: overall intercept & race). We fit separate models for each of the outcomes. All numerical predictors were standardized at the appropriate level (county, state) before model estimation (for post-stratification as well as for final inference) to help with estimation efficiency and interpretability. We also set priors for the intercept and coefficients in the bayesian model to be weakly informative normal distributions centered on zero with a standard deviation of five. This corresponds to a prior belief of all parameters taking values between -15 and 15. Values outside this range are extremely unlikely given that all variables were standardized prior to estimation. All other parameters were left to default values.

## Project Implicit Estimates
```{r exp-diff}
individual_data %>%
  mutate(explicit_bias_diff = twhite_0to10-tblack_0to10) -> individual_data
```
We first report the results of estimating the implicit and explicit biases in from Project Implicit data. When examining the unstandardized county-level estimates adjusted with poststratification, we see that there is a pro-white bias in both implicit ($mean$ = `r round(mean(as.numeric(county_means$weighted_bias), na.rm=T), 2)`, $sd$ = `r round(sd(as.numeric(county_means$weighted_bias), na.rm=T), 2)`) and explicit measures ($mean$ = `r round(mean(as.numeric(county_means$weighted_explicit_diff), na.rm=T), 2)`, $sd$ = `r round(sd(as.numeric(county_means$weighted_explicit_diff), na.rm=T), 2)`, where on both scales 0 = no bias, and positive numbers indicate a pro-white bias).

## Disciplinary action frequency

```{r disc-count, results='asis'}
model.data %>% 
  filter(exclude==FALSE) %>%
  group_by(group, metric) %>% 
  summarise(percentage = sum(number)/sum(total_number)) %>%
  mutate(percentage = percentage*100) %>%
  mutate(percentage = paste(round(percentage, 2), '%', sep='')) %>%
  spread(group, percentage) %>%
  mutate(metric = fct_recode(metric, 
                             'expulsions' = 'expulsion_combined',
                             'school arrests' = 'in_school_arrest',
                             'in-school suspension' = 'inschool_susp',
                             'law enforcement referral' = 'law_enforcement',
                             'out-of-school suspension' = 'oos_susp')) -> counts_table

papaja::apa_table(counts_table, caption = 'Percentage of students of each race receiving each type of disciplinary action')
```

Table \@ref(tab:disc-count) shows the percentage of students of each race who were reported having received each of the actions under consideration. Our statistical models (described below) show that these differences are extremely unlikely to be due to chance.

## Associations across counties
Figure \@ref(fig:overall-associations) shows the estimate of primary interest for each of the models. The estimates displayed are the coefficients for the interaction between race and each of the two bias measurements. Given that African Americans are the baseline group, negative values for this coefficient indicate that as one moves into counties with higher levels of bias, the gap between the probability of a black student being disciplined and the probability of a white student being disciplined grows. Table \@ref(tab:reg-coefs) displays the estimated coefficients and 95% uncertainty intervals for each of the fixed effects for each model.

```{r overall-associations, fig.cap="Association between each metric and county-level estimates of explicit and implicit bias. Negative values indicate that the rate of increase (or decrease) for blacks is faster (or slower) than for whites. Point is the mean of the posterior and error bars represent 95% bayesian uncertainty intervals."}
#the p_neg for oos_susp & implicit bias is 100%, so when reporting the number in-text, i just replace it with 1 to prevent it from being rendered as NA in the document.
base_path <- '/Users/travis/Documents/gits/educational_disparities/cluster/output/mw_uclaexcl_diff_extracovs/'
plot.dat <- data.frame(group = NA, weighted_bias = NA, est = NA, lower = NA, upper=NA, metric = NA)
plot.dat2 <- data.frame(weighted_bias = NA, est = NA, lower=NA, upper=NA, metric=NA)
plot.dat3 <- data.frame(est=NA, lower=NA, upper=NA, p_neg=NA, metric=NA)

# weighted bias w/exclusions & explicit difference
metrics <- list.files(base_path)
#metrics <- c('expulsion_combined', 'inschool_susp', 'oos_susp')
for (i in metrics){
  pth <- paste(base_path, i, '/', sep='')

  files <- list.files(pth)
  j <- 1
  posterior_combo <- array(0, dim=c(17,4000,length(files)))
  raw_rate <- data.frame(county_id=NA, nschools=NA, weighted_bias=NA, group=NA, nincidents=NA, nstudents=NA, rate=NA)
  for (k in files){
    #print(i)
    load(paste(pth,k,sep=''))
    dat <- as.matrix(m)
    m$data %>% 
      select(county_id, COMBOKEY) %>% 
      distinct() %>% 
      group_by(county_id) %>% 
      summarise(nschools=n()) -> n_schools
    m$data %>%
      group_by(county_id, group) %>%
      summarise(nincidents=sum(number), nstudents=sum(total_number)) %>%
      mutate(rate=nincidents/nstudents) -> rate
    n_schools %>%
      left_join(m$data[, c('county_id', 'weighted_bias')]) %>%
      distinct() %>%
      left_join(rate) %>%
      rbind(raw_rate) %>%
      filter(!is.na(county_id)) -> raw_rate
    posterior_combo[,,j] <- t(dat[,c(1:17)])
    j <- j+1
  }
  p_cons <- parallelMCMCcombine::consensusMCcov(posterior_combo, shuff = T)

  for (p in 1:17){
    m$stanfit@sim$samples[[1]][[p]] <- p_cons[p,1:1000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[2]][[p]] <- p_cons[p,1001:2000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[3]][[p]] <- p_cons[p,2001:3000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[4]][[p]] <- p_cons[p,3001:4000]
  }

  data.frame(est=mean(p_cons[16,]),
             lower=quantile(p_cons[16,], .025),
             upper=quantile(p_cons[16,], .975),
             p_neg = prop.table(table(p_cons[16,]<0))[2],
             metric=i) %>%
    rbind(plot.dat3) %>%
    filter(!is.na(est)) %>%
    mutate(metric=reorder(metric, est)) -> plot.dat3
}

plot.dat3$metric <- fct_recode(plot.dat3$metric, 
                               'Out-of-School Suspension' = 'oos_susp',
                               'In-School Suspension' = 'inschool_susp',
                               'School-Related Arrest' = 'in_school_arrest',
                               'Law Enf. Referral' = 'law_enforcement',
                               'Expulsion' = 'expulsion_combined')
plot.dat3_all <-plot.dat3
plot.dat3_all$bias <- 'Implicit'

#weighted warmth w/exclusions & explicit difference
plot.dat <- data.frame(group = NA, weighted_explicit_diff = NA, est = NA, lower = NA, upper=NA, metric = NA)
plot.dat2 <- data.frame(weighted_explicit_diff = NA, est = NA, lower=NA, upper=NA, metric=NA)
plot.dat3 <- data.frame(est=NA, lower=NA, upper=NA, p_neg = NA, metric=NA)

for(i in metrics){
  pth <- paste(base_path, i, '/', sep='')

  files <- list.files(pth)
  j <- 1
  posterior_combo <- array(0, dim=c(17,4000,length(files)))
  raw_rate_warmth <- data.frame(county_id=NA, nschools=NA, weighted_explicit_diff=NA, group=NA, nincidents=NA, nstudents=NA, rate=NA)
  for (k in files){
    #print(i)
    load(paste(pth,k,sep=''))
    dat <- as.matrix(m)
    m$data %>% 
      select(county_id, COMBOKEY) %>% 
      distinct() %>% 
      group_by(county_id) %>% 
      summarise(nschools=n()) -> n_schools
    m$data %>%
      group_by(county_id, group) %>%
      summarise(nincidents=sum(number), nstudents=sum(total_number)) %>%
      mutate(rate=nincidents/nstudents) -> rate
    n_schools %>%
      left_join(m$data[, c('county_id', 'weighted_explicit_diff')]) %>%
      distinct() %>%
      left_join(rate) %>%
      rbind(raw_rate_warmth) %>%
      filter(!is.na(county_id)) -> raw_rate_warmth
    posterior_combo[,,j] <- t(dat[,c(1:17)])
    j <- j+1
  }
  p_cons <- parallelMCMCcombine::consensusMCcov(posterior_combo, shuff = T)

  for (p in 1:17){
    m$stanfit@sim$samples[[1]][[p]] <- p_cons[p,1:1000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[2]][[p]] <- p_cons[p,1001:2000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[3]][[p]] <- p_cons[p,2001:3000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[4]][[p]] <- p_cons[p,3001:4000]
  }

  data.frame(est=mean(p_cons[17,]),
             lower=quantile(p_cons[17,], .025),
             upper=quantile(p_cons[17,], .975),
             p_neg = prop.table(table(p_cons[17,]<0))[2],
             metric=i) %>%
    rbind(plot.dat3) %>%
    filter(!is.na(est)) %>%
    mutate(metric=reorder(metric, est)) -> plot.dat3
}

plot.dat3$bias <- 'Explicit'
plot.dat3$metric <- fct_recode(plot.dat3$metric, 
                               'Out-of-School Suspension' = 'oos_susp',
                               'In-School Suspension' = 'inschool_susp',
                               'School-Related Arrest' = 'in_school_arrest',
                               'Law Enf. Referral' = 'law_enforcement',
                               'Expulsion' = 'expulsion_combined')
plot.dat3_all <- rbind(plot.dat3_all, plot.dat3)
ggplot(plot.dat3_all, aes(x=metric, y=est, group=bias, color=bias, shape=bias)) +
  geom_point(position=position_dodge(width=.5), size=2) +
  geom_errorbar(aes(ymin=lower, ymax=upper), 
                width=.5, position=position_dodge(width=.5)) +
  theme_classic() +
  coord_flip() +
  geom_hline(yintercept=0) +
  ylab('Black-White Difference in log-odds slope') +
  theme(axis.title.y = element_blank(),
        legend.title = element_blank())
```

Several patterns are apparent from this figure. First, in all cases but school-related arrests, county-level explicit biases are reliably associated with disciplinary disparities, such that counties with more bias are expected to have a larger racial gap. The effect with the largest magnitude of these is out-of-school suspensions ($M_{slope}$ = `r round(plot.dat3_all$est[6], 2)`, [`r round(plot.dat3_all$lower[6], 2)`, `r round(plot.dat3_all$upper[6], 2)`]), followed by expulsions ($M_{slope}$ = `r round(plot.dat3_all$est[10], 2)`, [`r round(plot.dat3_all$lower[10], 2)`, `r round(plot.dat3_all$upper[10], 2)`]), law enforcement referrals ($M_{slope}$ = `r round(plot.dat3_all$est[7], 2)`, [`r round(plot.dat3_all$lower[7], 2)`, `r round(plot.dat3_all$upper[7], 2)`]), and in-school suspensions ($M_{slope}$ = `r round(plot.dat3_all$est[8], 2)`, [`r round(plot.dat3_all$lower[8], 2)`, `r round(plot.dat3_all$upper[8], 2)`]). The model indicates that the evidence for school-related arrests is inconclusive ($M_{slope}$ = `r round(plot.dat3_all$est[9], 2)`, [`r round(plot.dat3_all$lower[9], 2)`, `r round(plot.dat3_all$upper[9], 2)`]).

Second, the effect with the largest magnitude of all is that between implicit biases and out-of-school suspension ($M_{slope}$ = `r round(plot.dat3_all$est[1], 2)`, [`r round(plot.dat3_all$lower[1], 2)`, `r round(plot.dat3_all$upper[1], 2)`]). All other estimated parameters are inconclusive.

```{r, reg-coefs, results='asis'}
base_path <- '/Users/travis/Documents/gits/educational_disparities/cluster/output/mw_uclaexcl_diff_extracovs/'
plot.dat <- data.frame(coef = NA, est = NA, lower = NA, upper = NA, p_neg = NA, metric = NA)

# weighted bias w/exclusions & explicit difference
metrics <- list.files(base_path)
#metrics <- c('expulsion_combined', 'inschool_susp', 'oos_susp')
for (i in metrics){
  pth <- paste(base_path, i, '/', sep='')

  files <- list.files(pth)
  j <- 1
  posterior_combo <- array(0, dim=c(17,4000,length(files)))
  for (k in files){
    #print(i)
    load(paste(pth,k,sep=''))
    dat <- as.matrix(m)
    posterior_combo[,,j] <- t(dat[,c(1:17)])
    j <- j+1
  }
  p_cons <- parallelMCMCcombine::consensusMCcov(posterior_combo, shuff = T)

  for (p in 1:17){
    m$stanfit@sim$samples[[1]][[p]] <- p_cons[p,1:1000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[2]][[p]] <- p_cons[p,1001:2000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[3]][[p]] <- p_cons[p,2001:3000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[4]][[p]] <- p_cons[p,3001:4000]
  }

  data.frame(coef=colnames(dat)[1:17],
             est=rowMeans(p_cons),
             lower=apply(p_cons,1,quantile, probs=.025),
             upper=apply(p_cons,1,quantile, probs=.975),
             p_neg = rowSums(p_cons<0)/4000,
             metric=rep(i, 17)) %>%
    rbind(plot.dat) %>%
    filter(!is.na(est)) -> plot.dat
}

plot.dat$metric <- fct_recode(plot.dat$metric, 
                               'Out-of-School Suspension' = 'oos_susp',
                               'In-School Suspension' = 'inschool_susp',
                               'School-Related Arrest' = 'in_school_arrest',
                               'Law Enf. Referral' = 'law_enforcement',
                               'Expulsion' = 'expulsion_combined')
plot.dat$coef <- fct_recode(plot.dat$coef,
                            'Intercept' = '(Intercept)',
                            'total population' = 'total_pop',
                            'propotion black' = 'black_prop',
                            'proportion white' = 'white_prop',
                            'black-white ratio' = 'b.w.ratio',
                            'college grads' = 'col_grads',
                            'income' = 'med_income',
                            'poverty' = 'poverty_rate',
                            'unemployment' = 'unemp_rate',
                            'crime' = 'crime_rate',
                            'housing density' = 'density',
                            'mobility' = 'mobility',
                            'race: white' = 'groupwhite',
                            'implicit bias' = 'weighted_bias',
                            'explicit bias' = 'weighted_explicit_diff',
                            'implicit bias*race: white' = 'groupwhite:weighted_bias',
                            'explicit bias*race: white' = 'groupwhite:weighted_explicit_diff')

plot.dat %>%
  mutate(estimate=paste(round(est, 2), ', [', round(lower, 2), ',', round(upper, 2),']', sep='')) %>%
  select(metric, estimate, coef) %>%
  spread(metric, estimate) -> tabledata

tabledata <- tabledata[c(1,13,3,17,2,4,10,12,14,5,6,11,7,15,16,8,9),]
names(tabledata)[1] <- ''

papaja::apa_table(tabledata, caption = 'Regression coefficient estimates for the population-level (i.e. fixed) effects, along with 95\\% uncertainty intervals for each of the disciplinary metrics.', small = T)
```

The findings for other disciplinary metrics suggest a differential effect of explicit and implicit bias. For example, there is a reliable association between in-school suspensions and explicit bias. The difference in the slope of the association between explicit bias and the log of the odds for in-school suspensions between white and black students is estimated to be `r round(plot.dat3_all$est[6], 2)`, with 95% of the posterior distribution between `r round(plot.dat3_all$lower[6], 2)` and `r round(plot.dat3_all$upper[6], 2)` and a proportion `r afex::round_ps(1)` of the posterior distribution consistent with a negative effect. Similar effects can be seen for explicit bias and law enforcement referrals (`r round(plot.dat3_all$est[7], 2)`, [`r round(plot.dat3_all$lower[7],2)`, `r round(plot.dat3_all$upper[7],2)`], $p_{neg}$ `r afex::round_ps(plot.dat3_all$p_neg[7])`), and explicit bias and expulsions (`r round(plot.dat3_all$est[10], 2)`, [`r round(plot.dat3_all$lower[10],2)`, `r round(plot.dat3_all$upper[10],2)`], $p_{neg}$ = `r afex::round_ps(plot.dat3_all$p_neg[10])`). In contrast, zero is included as a plausible value for the associations between implicit bias and in-school suspensions (`r round(plot.dat3_all$est[3], 2)`, [`r round(plot.dat3_all$lower[3],2)`, `r round(plot.dat3_all$upper[3],2)`], $p_{neg}$ = `r afex::round_ps(plot.dat3_all$p_neg[3])`), law enforcement referrals (`r round(plot.dat3_all$est[2], 2)`, [`r round(plot.dat3_all$lower[2],2)`, `r round(plot.dat3_all$upper[2],2)`], $p_{neg}$ = `r afex::round_ps(plot.dat3_all$p_neg[2])`), and expulsions (`r round(plot.dat3_all$est[5], 2)`, [`r round(plot.dat3_all$lower[5],2)`, `r round(plot.dat3_all$upper[5],2)`], $p_{neg}$ = `r afex::round_ps(plot.dat3_all$p_neg[5])`).

```{r detail-risks, fig.cap="Association between bias and relative risk ratio for black students to white students. Line is the mean of the posterior. Bands indicate 95% uncertainty intervals. Points represent counties, whose sizes are scaled to the number of schools in that county. Note that the y axis is cut at 8 for legibility despite data extending beyond."}
plot.dat <- data.frame(group=NA, weighted_bias=NA, est=NA, 
                       lower=NA, upper=NA, bias=NA, metric=NA)
plot.dat2 <- data.frame(weighted_bias=NA, est=NA, 
                        lower=NA, upper=NA, bias=NA, metric=NA)
raw_or <- data.frame(county_id=NA, nschools=NA, bias=NA, 
                     value=NA, black=NA, white=NA, odds_r=NA, metric=NA)
for (i in metrics){
  pth <- paste(base_path, i, '/', sep='')

  files <- list.files(pth)
  j <- 1
  posterior_combo <- array(0, dim=c(17,4000,length(files)))
  for (k in files){
  raw_rate <- data.frame(county_id=NA, nschools=NA, weighted_bias=NA, 
                         weighted_explicit_diff=NA, group=NA, nincidents=NA, 
                         nstudents=NA, rate=NA)
    #print(i)
    load(paste(pth,k,sep=''))
    df <- as.matrix(m)
    m$data %>% 
      select(county_id, COMBOKEY) %>% 
      distinct() %>% 
      group_by(county_id) %>% 
      summarise(nschools=n()) -> n_schools
    m$data %>%
      group_by(county_id, group) %>%
      summarise(nincidents=sum(number), nstudents=sum(total_number)) %>%
      mutate(rate=nincidents/nstudents) -> rate
    n_schools %>%
      left_join(m$data[, c('county_id', 'weighted_bias', 'weighted_explicit_diff')]) %>%
      distinct() %>%
      left_join(rate) %>%
      rbind(raw_rate) %>%
      filter(!is.na(county_id)) -> raw_rate
    posterior_combo[,,j] <- t(df[,c(1:17)])
    j <- j+1
  }
  p_cons <- parallelMCMCcombine::consensusMCcov(posterior_combo, shuff = T)

  for (p in 1:17){
    m$stanfit@sim$samples[[1]][[p]] <- p_cons[p,1:1000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[2]][[p]] <- p_cons[p,1001:2000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[3]][[p]] <- p_cons[p,2001:3000]
  }
  for (p in 1:17){
    m$stanfit@sim$samples[[4]][[p]] <- p_cons[p,3001:4000]
  }

  newdat <- expand.grid(group = rownames(table(m$data$group)),
                        total_pop=0,
                        unemp_rate=0,
                        med_income=0,
                        poverty_rate=0,
                        col_grads=0,
                        white_prop=0,
                        black_prop=0,
                        b.w.ratio=0,
                        mobility=0,
                        crime_rate=0,
                        density=0,
                        weighted_bias = seq(-2.5,2.5,.25),
                        weighted_explicit_diff = 0,
                        number=0,
                        total_number=1)
  y_hat <- rstanarm::posterior_linpred(m, newdata = newdat, re.form=~0, transform=T)

  cbind(newdat, t(y_hat)) %>% 
    mutate(index=row_number()) %>%
    gather(sample, value, `1`:`4000`) -> posterior_distribution

  posterior_distribution %>%
    select(group, weighted_bias, sample, value, index) %>%
    group_by(group, weighted_bias) %>%
    summarise(est = mean(value),
              lower = quantile(value, .025),
              upper = quantile(value, .975)) %>%
    ungroup() %>%
    mutate(bias = 'Implicit',
           metric = i) %>%
    rbind(plot.dat) -> plot.dat
